---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(scipen = 999)

```

# dQALY
<!-- badges: start -->
<!-- badges: end -->

<span style="color:red"> ***This package is currently under active development and the code is subject to change.***  </span>


The goal of the dQALY package is to provide an easy and flexible way of calculating the number of QALYs that are 'lost' when a person dies. 

This package has been built using code adapted from the [COVID19_QALY_App](https://github.com/LSHTM-GHECO/COVID19_QALY_App), written by Nichola Naylor at LSHTM. The app was itself adapted from an [Excel tool](https://avalonecon.com/estimating-qaly-losses-associated-with-deaths-in-hospital-covid-19/) built by Andrew Briggs to operationalise the methods he & others set out in a [letter](https://onlinelibrary.wiley.com/doi/10.1002/hec.4208) published in the journal Health Economics in 2020. 


## Installation

You can install the development version of dQALY from [GitHub](https://github.com/) with:

```{r, message=FALSE}
# install.packages("pak")
pak::pak("katehayes/dQALY")
```

## Calculating QALY loss due to death - get started

```{r first_example}
library(dQALY)

# you have to minimally specify country and year
calculate_dQALY(country = "United Kingdom", year = 2015)

# Note: If country or year you choose aren't available, it will tell you
# calculate_dQALY(country = "Scotland", year = 2015)
# calculate_dQALY(country = "United Kingdom", year = 2010)


# You can also output mean dQALY values for population groups:
# Outputting one average value for both sexes together
# calculate_dQALY(country = "United Kingdom", year = 2015, sex_group = T)

# Outputting average values for a set of age groups you specify
# my_age_groups <- data.table(age_low = c(seq(0,90,5)), age_high = c(seq(4,89,5), 100))
# calculate_dQALY(country = "United Kingdom", year = 2015, age_groups = my_age_groups)

# Outputting average values for both sexes and for the specified age groups
# calculate_dQALY(country = "United Kingdom", year = 2015, sex_group = T, age_groups = my_age_groups)

```

## Population-average health utility scores (utility norms)

The available utility norms can be viewed using the `get_norm_info` function. Results can be filtered by country, and returned with or without reference information.

```{r get_norm_info}

# Return all utility norm sets for all countries (10 rows only)
print(get_norm_info(), class = FALSE, nrows = 10)

# Return all English utility norm sets with reference information
# print(get_norm_info(country = "England", references = T), class = FALSE)

```

For some countries, there is more than one set of utility norms stored in the package data. For every country, we have chosen a set of default norms - these are the norms that will be used when the `calculate_dQALY` function is called without specifying a value for the `norms` argument. Alternatively, the user specify what set of package norms they would like to use by passing its ID to `norms`. Its also possible to supply custom norms.

```{r package_norm_plots}

library(ggplot2)

# Calculating dQALY for four different countries using the country-specific 
# default utility norms set by the package
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              sex_group = T),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   sex_group = T),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   sex_group = T),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   sex_group = T),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


# Calculating dQALY using the same set of utility norms for every country
# note that the ID used needs to be valid. e.g.:
# calculate_dQALY(country = "England", year = 2019, norms = "invalid") won't work
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              sex_group = T,
                              norms = "janssen_euvas"),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   sex_group = T,
                                   norms = "janssen_euvas"),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   sex_group = T,
                                   norms = "janssen_euvas"),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   sex_group = T,
                                   #in hungary's case, EUVAS norms
                                   #are already the default
                                   norms = "janssen_euvas"),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()



```
```{r custom_norms}
library(data.table)

# how to specify your own set of utility norms
my_norms <- data.table(sex = c(rep("male", 3), 
                               rep("female", 3)),
                       age_low = c(0, 20, 90),
                       age_high = c(19, 89, 150),
                       avg_util = c(1, 0.85, 0.67,
                                    0.99, 0.4, 0.2))

# if we set the health utility score for all ages equal to 1
# then we'll be calculating life years as opposed to QALYs lost
no_quality_adjustment <- data.table(sex = c("male", "female"),
                                    age_low = 0,
                                    age_high = 150,
                                    avg_util = 1)


ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              sex_group = T),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "black") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   norms = no_quality_adjustment),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()




```

## Discounting
To get the net present value of the losses, we apply a discount rate. In our package, the default discount rate is set at 3.5% as per the [NICE health technology evaluations manual](https://www.nice.org.uk/process/pmg36/chapter/economic-evaluation-2#discounting). However, the most appropriate discount rate to apply in a given evaluation/analysis is commonly subject to debate. [The Green Book](https://www.gov.uk/government/publications/the-green-book-appraisal-and-evaluation-in-central-government/the-green-book-2020#a6-discounting), guidance on evaluation methods issued by the Treasury, discusses a number of discounting regimes and the reasons one might use them. 

Our package allows discount rates to be specified flexibly.

```{r discounting}

# Defining a number of different discount regimes
# No discounting
r_none <- 0 #equivalently, r_none <- function(x) 0

# NICE reference case discount rate/ Green Book standard Social Time Preference Rate
r_standard <- 0.035 #equivalently, r_standard <- function(x) 0.035
# NICE alternative discount rate/Green Book recommended discount rate for health or life values
r_health <- 0.015 #equivalently, r_health <- function(x) 0.015

# Long term discounting
# Green Book recommended declining long term discount rate for health or life values
r_health_lt <- function(x) ifelse(x < 31, 0.015, ifelse(x > 75, 0.0107, 0.0129))
# Green Book recommended rate reduced by excluding pure social time preference
# (relevant if intervention may effect substantial/irreversible wealth transfers between generations)
r_health_lt_reduced <- function(x) ifelse(x < 31, 0.01, ifelse(x > 75, 0.0071, 0.0086))



# Plotting QALY loss due to death for England in 2019, 
# as discount regimes vary
library(dQALY)
library(ggplot2)

ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              sex_group = T), #equivalently, r = r_standard
       aes(x = age_at_death, y = dQALY)) +
  geom_line() +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health_lt),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health_lt_reduced),
            colour = "purple") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_none),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 80),
                     expand = c(0,0)) +
  theme_classic()


```


