---
output: github_document
---
<!-- devtools::build_readme() -->
<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# options(scipen = 999)

```

# dQALY
<!-- badges: start -->
<!-- badges: end -->

<span style="color:red"> ***This package is currently under active development and the code is subject to change.***  </span>

The quality-adjusted life year, or QALY, is a widely used outcome measure in the field of health economics. 
When evaluating the impact of a policy/programme/intervention, we often want to express the health impacts of preventing or failing to prevent a death in terms of the QALYs that would be gained or lost.

The goal of the dQALY package is to provide an easy and flexible way of calculating the number of QALYs that are 'lost' when a person dies. 

This package has been built using code adapted from the [COVID19_QALY_App](https://github.com/LSHTM-GHECO/COVID19_QALY_App), written by Nichola Naylor at LSHTM. The [COVID19_QALY_App](https://github.com/LSHTM-GHECO/COVID19_QALY_App) was itself adapted from an [Excel tool](https://avalonecon.com/estimating-qaly-losses-associated-with-deaths-in-hospital-covid-19/) built by Andrew Briggs to operationalise the methods for calculating QALY loss due to death that he & others set out in a [letter](https://onlinelibrary.wiley.com/doi/10.1002/hec.4208) published in the journal Health Economics in 2020.
<!-- Some of the setup borrowed from qalytools & eq5d packages -->



## Installation

You can install the development version of dQALY from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("pak")
pak::pak("katehayes/dQALY")
```

## Calculating QALY loss due to death with life expectancy & quality of life data

Add overview of methods. Using life tables & utility norms. Utility norms themselves are constructed from health state data & value sets. 

## Population-average health related quality of life scores (utility norms)

Some discussion of utility norms can be found on the [EuroQol website](https://euroqol.org/information-and-support/resources/population-norms/):

> Population reference data, sometimes called population norm data or simply population norms, provide baseline or reference scores for a specific country or international region, based on representative samples. EQ-5D population norms can be used to to benchmark the outcomes of specific individuals or groups (e.g. patients) against the health of the general population, which also helps to identity the burden of the disease of patients or patient groups.

For most countries, there is more than one set of utility norms stored in the package data. For every country, we have chosen a set of default norms - these are the norms that will be used when the `calculate_dQALY` function is called without specifying a value for the `norms` argument. Alternatively, the user can specify what set of package norms they would like to use by passing its ID to `norms`. Its also possible to supply custom norms.

The list of available utility norms and their IDs can be viewed using the `get_norm_info` function. This function also returns information we have documented about how each set of utility norms - specifically, about the EQ-5D data and value sets from which the norms are estimated. We have tried to adopt the same terminology/categorisation scheme used by the eq5d package to document information about value sets. 
<!-- Note: we could also store info about the model used to estimate the pop-level norms from the input data (eq5d profiles valued w the value sets) - e.g. in the ViH paper they use a linear model -->

Having this information about the make-up of the utility norms is useful - information about, for example, what population the health state was data gathered from/ what population valued the health states/ when the was data collected/ what methods were used to elicit the states/values. Being able to access this information allows the user to understand the implications of choosing one set of norms over another - and to make a judgement about the set of norms that is most appropriate for their purposes. 
<!-- What modelling methods were used to estimate population averages? -->

## Discounting
To get the net present value of the losses, we apply a discount rate. In our package, the default discount rate is set at 3.5% as per the [NICE health technology evaluations manual](https://www.nice.org.uk/process/pmg36/chapter/economic-evaluation-2#discounting). However, the most appropriate discount rate to apply in a given evaluation/analysis is commonly subject to debate. [The Green Book](https://www.gov.uk/government/publications/the-green-book-appraisal-and-evaluation-in-central-government/the-green-book-2020#a6-discounting), guidance on evaluation methods issued by the Treasury, discusses a number of discounting regimes and the reasons one might use them. 

Our package allows discount rates to be specified flexibly.




<table>
    <thead>
        <tr>
            <th>Command</th>
            <th>Explanation</th>
            <th>Years 0 – 30</th>
            <th>Years 31 – 75</th>
            <th>Years 76 – 125</th>
        </tr>
    </thead>
    <tbody>
        <tr>
            <td>r_none()</td>
            <td>No discounting</td>
            <td>0%</td>
            <td>0%</td>
            <td>0%</td>
        </tr>
        <tr>
            <td>r_default()</td>
            <td>NICE reference case discount rate/ Green Book standard Social Time Preference Rate</td>
            <td>3.5%</td>
            <td>3.5%</td>
            <td>3.5%</td>
        </tr>
        <tr>
            <td>r_health()</td>
            <td>NICE alternative discount rate/Green Book recommended discount rate for health or life values</td>
            <td>1.5%</td>
            <td>1.5%</td>
            <td>1.5%</td>
        </tr>
        <tr>
            <td>r_lt_health()</td>
            <td>Green Book recommended declining long term discount rate for health or life values</td>
            <td>1.5%</td>
            <td>1.29%</td>
            <td>1.07%</td>
        </tr>
        <tr>
            <td>r_lt_health_reduced()</td>
            <td>Green Book recommended rate reduced by excluding pure social time preference (relevant if intervention may effect substantial/irreversible wealth transfers between generations)</td>
            <td>1%</td>
            <td>0.86%</td>
            <td>0.71%</td>
        </tr>
    </tbody>
</table>
















