---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

options(scipen = 999)
```

# dQALY
<!-- badges: start -->
<!-- badges: end -->

<span style="color:red"> ***This package is currently under active development and the code is subject to change.***  </span>

Our package was originally adapted from the [COVID19_QALY_App](https://github.com/LSHTM-GHECO/COVID19_QALY_App) built by Nichola Naylor at LSHTM. This app was itself adapted from an [Excel tool](https://avalonecon.com/estimating-qaly-losses-associated-with-deaths-in-hospital-covid-19/) built by Andrew Briggs to operationalise the methods he & others set out a [letter](https://onlinelibrary.wiley.com/doi/10.1002/hec.4208) published in the journal Health Economics in 2020. 

The goal of dQALY is to ...

## Installation

You can install the development version of dQALY from [GitHub](https://github.com/) with:

```{r, message=FALSE}
# install.packages("pak")
pak::pak("katehayes/dQALY")
```


```{r example_calculation}

library(dQALY)

calculate_dQALY(country = "England", year = 2019)

```



## Discounting
To get the net present value of the losses, we apply a discount rate. Our default discount rate is set at 3.5% as per the [NICE health technology evaluations manual](https://www.nice.org.uk/process/pmg36/chapter/economic-evaluation-2#discounting). However, the most appropriate discount rate to apply in a given evaluation/analysis is commonly subject to debate. [The Green Book](https://www.gov.uk/government/publications/the-green-book-appraisal-and-evaluation-in-central-government/the-green-book-2020#a6-discounting), guidance on evaluation methods issued by the Treasury, discusses a number of discounting regimes and the reasons one might use them. 

Our package allows discount rates to be specified flexibly.

```{r discounting}

# Defining a number of different discount regimes
# No discounting
r_none <- 0 #equivalently, r_none <- function(x) 0

# NICE reference case discount rate/ Green Book standard Social Time Preference Rate
r_standard <- 0.035 #equivalently, r_standard <- function(x) 0.035
# NICE alternative discount rate/Green Book recommended discount rate for health or life values
r_health <- 0.015 #equivalently, r_health <- function(x) 0.015

# Long term discounting
# Green Book recommended declining long term discount rate for health or life values
r_health_lt <- function(x) ifelse(x < 31, 0.015, ifelse(x > 75, 0.0107, 0.0129))
# Green Book recommended rate reduced by excluding pure social time preference
# (relevant if intervention may effect substantial/irreversible wealth transfers between generations)
r_health_lt_reduced <- function(x) ifelse(x < 31, 0.01, ifelse(x > 75, 0.0071, 0.0086))



# Plotting QALY loss due to death for England in 2019, 
# as discount regimes vary

library(ggplot2)

ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              sex_group = T), #equivalently, r = r_standard
       aes(x = age_at_death, y = dQALY)) +
  geom_line() +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health_lt),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_health_lt_reduced),
            colour = "purple") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   sex_group = T,
                                   r = r_none),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 80),
                     expand = c(0,0)) +
  theme_classic()

  


```


## Utility norms

The available utility norms can be viewed using the ***get_norm_info*** function. Results can be filtered by country, and returned with or without reference information.

```{r get_norm_info}


# Return all utility norm sets for all countries (10 rows only)
print(get_norm_info(), class = FALSE, nrows = 10)

# Return all English utility norm sets with reference information
print(get_norm_info(country = "England", references = T), class = FALSE)

```

