---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dQALY
<!-- badges: start -->
<!-- badges: end -->

<span style="color:red"> ***This package is currently under active development and the code subject to change.***  </span>

Our package was originally adapted from the [COVID19_QALY_App](https://github.com/LSHTM-GHECO/COVID19_QALY_App) built by Nichola Naylor at LSHTM. This app was itself adapted from an [Excel tool](https://avalonecon.com/estimating-qaly-losses-associated-with-deaths-in-hospital-covid-19/) built by Andrew Briggs to operationalise the methods he & others set out a [letter](https://onlinelibrary.wiley.com/doi/10.1002/hec.4208) published in the journal Health Economics in 2020. 

The goal of dQALY is to ...

## Installation

You can install the development version of dQALY from [GitHub](https://github.com/) with:

```{r, message=FALSE}
# install.packages("pak")
pak::pak("katehayes/dQALY")
```

## Utility Norms

The available utility norms can be viewed using the ***get_norm_info*** function. Results can be filtered by country, and returned with or without reference information.

```{r get_norm_info}
library(dQALY)

# Return all utility norm sets for all countries (first 6 only)
head(get_norm_info())

# Return all English utility norm sets with reference information
get_norm_info(country = "England", references = T)

```

