---
title: "Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dQALY)
```

## Calculating QALY loss due to death - get started

```{r first_example}

# you have to minimally specify country and year
# calculate_dQALY(country = "United Kingdom", year = 2015)

# Note: If country or year you choose aren't available, it will tell you
# calculate_dQALY(country = "Scotland", year = 2015)
# calculate_dQALY(country = "United Kingdom", year = 2010)


# You can also output mean dQALY values for population groups:
# Outputting one average value for both sexes together
# calculate_dQALY(country = "United Kingdom", year = 2015, collapse_sex = T)

# Outputting average values for a set of age groups you specify
# my_age_groups <- data.table(lower = 90, upper = 100)
# calculate_dQALY(country = "United Kingdom", year = 2015, collapse_age = my_age_groups)

# Outputting average values for both sexes and for the specified age groups
# calculate_dQALY(country = "United Kingdom", year = 2015, collapse_sex = T, collapse_age = my_age_groups)

```

## Population-average health related quality of life scores (utility norms)

For most countries, there is more than one set of utility norms stored in the package data. For every country, we have chosen a set of default norms - these are the norms that will be used when the `calculate_dQALY` function is called without specifying a value for the `norms` argument. Alternatively, the user specify what set of package norms they would like to use by passing its ID to `norms`. Its also possible to supply custom norms.

The available utility norms can be viewed using the `hrqol_norms` function. Results can be filtered by country, and returned with or without reference information.

```{r hrqol_norms}

# Return all utility norm sets for all countries
head(hrqol_norms())

# Return all English utility norm sets with reference information
head(hrqol_norms(country = "England", references = T))

```



```{r package_norm_plots}

library(ggplot2)

# Calculating dQALY for four different countries using the country-specific 
# default utility norms set by the package
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   collapse_sex = T),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   collapse_sex = T),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   collapse_sex = T),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


# Calculating dQALY using the same set of utility norms for every country
# note that the ID used needs to be valid. e.g.:
# calculate_dQALY(country = "England", year = 2019, norms = "invalid") won't work
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T,
                              norms = "janssen_euvas"),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   collapse_sex = T,
                                   norms = "janssen_euvas"),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   collapse_sex = T,
                                   norms = "janssen_euvas"),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   collapse_sex = T,
                                   #in hungary's case, EUVAS norms
                                   #are already the default
                                   norms = "janssen_euvas"),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


```


```{r custom_norms}
library(data.table)

# how to specify your own set of utility norms
my_norms <- data.table(sex = c(rep("male", 3), 
                               rep("female", 3)),
                       lower = c(0, 20, 90),
                       upper = c(19, 89, 150),
                       avg_util = c(1, 0.85, 0.67,
                                    0.99, 0.4, 0.2))


ggplot(data = calculate_dQALY(country = "England", year = 2019,
                              collapse_sex = T),
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "black") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019,
                                   collapse_sex = T,
                                   norms = my_norms),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


```

## Discounting


```{r discounting}

# Defining a number of different discount regimes
# No discounting
r_none <- 0 #equivalently, r_none <- function(x) 0

# NICE reference case discount rate/ Green Book standard Social Time Preference Rate
r_standard <- 0.035 #equivalently, r_standard <- function(x) 0.035
# NICE alternative discount rate/Green Book recommended discount rate for health or life values
r_health <- 0.015 #equivalently, r_health <- function(x) 0.015

# Long term discounting
# Green Book recommended declining long term discount rate for health or life values
r_health_lt <- function(x) ifelse(x < 31, 0.015, ifelse(x > 75, 0.0107, 0.0129))
# Green Book recommended rate reduced by excluding pure social time preference
# (relevant if intervention may effect substantial/irreversible wealth transfers between generations)
r_health_lt_reduced <- function(x) ifelse(x < 31, 0.01, ifelse(x > 75, 0.0071, 0.0086))



# Plotting QALY loss due to death for England in 2019, 
# as discount regimes vary

ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T), #equivalently, r = r_standard
       aes(x = age_at_death, y = dQALY)) +
  geom_line(colour = "black") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   collapse_sex = T,
                                   r = r_health),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   collapse_sex = T,
                                   r = r_health_lt),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   collapse_sex = T,
                                   r = r_health_lt_reduced),
            colour = "purple") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019,
                                   collapse_sex = T,
                                   r = r_none),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 80),
                     expand = c(0,0)) +
  theme_classic()


```

## Note on methods
QALY loss at age x without discounting is equivalent to QALE at age x. QALY loss at age x without discounting or adjusting by health-related quality of life is life expectancy at age x.


```{r equivalencies}

# if we set the discount rate equal to zero, we'll be returning estimates of quality adjusted life expectancy
no_discounting <- 0

# if we ALSO set the health utility score for all ages equal to 1, we'll be returning life expectancy data
no_quality_adjustment <- data.table(sex = c("male", "female"),
                                    lower = 0,
                                    upper = 150,
                                    avg_util = 1)


ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T), #default norms, default r
       aes(x = age_at_death, y = dQALY)) +
  geom_line() +
  geom_line(data = calculate_dQALY(country = "England", year = 2019,
                                   collapse_sex = T,
                                   r = no_discounting),
            colour = "blue") +
  # geom_line(data = calculate_QALE(country = "England", year = 2019,
  #                                   collapse_sex = T),
  #            colour = "green") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   collapse_sex = T,
                                   r = no_discounting,
                                   norms = no_quality_adjustment),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 90),
                     expand = c(0,0)) +
  theme_classic()

```

## Worked example: using the dQALY package to value outputs of an infectious disease model
Imagine that we had built an infectious disease model, simulating CPE infections among hospital inpatients & the deaths resulting from those infections. Imagine that because of data availability, the model population does not have sex/age structure. We want to produce an estimate of the number of QALYs that would be lost on the death of an average hospital inpatient - this single value will be applied to all deaths. 

We'll use this example to demonstrate why one might use several of the features of the `calculate_dQALY` function - specifically, the features that enable the estimation of QALY losses for user-supplied cohorts and groups.


```{r worked_example_cohort}


# # Setting my cohort - all hospital inpatients
# # Using data on Hospital Admitted Patient Care Activity, 2019-20
# # https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity/2019-20/summary-reports---apc---patient
# 
# temp <- tempfile(fileext = ".xlsx")
# download.file(url = "https://digital.nhs.uk/binaries/content/documents/corporate-website/publication-system/statistical/hospital-admitted-patient-care-activity/2019-20/summary-reports---apc---patient/summary-reports---apc---patient/publicationsystem%3AbodySections%5B2%5D/publicationsystem%3AdataFile",
#               temp, mode = "wb")
# 
# hospital_cohort <- as.data.table(readxl::read_xlsx(temp, sheet = 1)) |> 
#   setnames(new = c("x", "male", "female")) |> 
#   melt(measure.vars = c("male", "female"),
#        variable.name = "sex",
#        value.name = "count")
# 
# hospital_cohort <- hospital_cohort[x != "Unknown"][
#   , lower := as.numeric(substring(x, 1, 2))
# ][
#   , upper := lower + 4
# ][
#   , .(x = c(lower:upper), count = rep(count/5, 5)), by = c("lower", "upper", "sex")
# ][
#   , c("lower", "upper"):=NULL
# ]
# 
# 
# # Setting my age groups - want to group all ages together
# all_ages <- data.table(lower = 0, upper = 94)
# 
# 
# # Getting my QALY loss estimate
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = hospital_cohort)
# 
# 
# 
# # What if we were only modelling adult inpatients?
# adult_hospital_cohort <- hospital_cohort[x >= 18]
# 
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = adult_hospital_cohort)
# 
# 
# 
# # What if mortality rates were particularly high among newborns?
# adj_hospital_cohort <- copy(hospital_cohort)[x == 0, count := count*3]
# 
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = adj_hospital_cohort)
# 
# 
# # What if we wanted to account for the fact that mortality and morbidity among
# # hospital patients is higher than it is among the general population?
# calculate_dQALY(country = "England", year = 2020,
#                 smr = 1.05, qcm = 0.95,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = hospital_cohort)
# # Note: if you had your own life expectancy or health utility data for a hospital population
# # you could supply it directly to the function
# 
# 
# # Examining my estimates
# ggplot(data = calculate_dQALY(country = "England", year = 2020, 
#                               collapse_sex = T),
#        aes(x = age_at_death, y = dQALY)) +
#   geom_line(colour = "black") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = hospital_cohort)$dQALY, 
#             colour = "red",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = adult_hospital_cohort)$dQALY, 
#             colour = "blue",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = adj_hospital_cohort)$dQALY, 
#             colour = "green",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           smr = 1.05, qcm = 0.95,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = hospital_cohort)$dQALY, 
#             colour = "purple",
#             linetype = "dashed") +
#   scale_x_continuous(name = "Age at death",
#                      limits = c(0, 125),
#                      expand = c(0,0)) +
#   scale_y_continuous(name = "QALY loss",
#                      limits = c(0, 30),
#                      expand = c(0,0)) +
#   theme_classic()

```

