---
title: "Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dQALY)
library(ggplot2)
library(data.table)
```


<!-- The below text is clipped from a description of a tool that researchers built. Useful articulation - can describe flexibility of dQALY package in similar manner -->

<!-- The disease models and default parameters implemented in the present version of the BCoDE toolkit represent the result of a thorough reviewing and validation process that was performed over a time period of around 5 years involving disease-specific experts from many countries and institutions. -->

<!-- However, since disease progression can be different in different populations (due to e.g. health care system, access to health care, social economic factors) and since default values within the toolkit outcome trees were chosen with the aim to reflect a European average, the software application enables the user to choose other than default parameter values and to include uncertainty in those parameters into the analysis. All disease model parameters can be thus edited by the user (see below). -->

<!-- The user become then responsible for the correctness of the assumptions made and of the validation. In fact, although we see this flexibility as a great advantage of this tool for carrying out tailored research, simulation exercises, scenario analyses and more accurate national and regional disease burden estimations taking into account geographical specificities and other peculiarities of different populations, we also want to emphasize the importance for the user of always choosing parameters based on sound clinical and epidemiological evidence for the validity of the burden estimation. -->









## Calculating QALY loss due to death - get started

You have to minimally specify country and year:
```{r}
calculate_dQALY(country = "United Kingdom", year = 2015) |> head()
calculate_dQALY(country = "France", year = 2020) |> head()
```
If country or year you choose aren't available, you'll get an error message informing you of that:
```{r, error = TRUE}
calculate_dQALY(country = "Scotland", year = 2015)
calculate_dQALY(country = "United Kingdom", year = 2010)
```
You can also output mean dQALY values for population groups. Here, we're outputting one average value for each year of age for both sexes together:
```{r}
calculate_dQALY(country = "United Kingdom", year = 2015, collapse_sex = T) |> head()
```
Here we're outputting average values for a set of age groups that we specify:
```{r}
my_age_groups <- data.table(lower = c(0, 90), upper = c(89, 99))
calculate_dQALY(country = "United Kingdom", year = 2015, collapse_age = my_age_groups)
```
And lastly, here we're outputting average values for both sexes together and for the specified age groups:
```{r}
calculate_dQALY(country = "United Kingdom", year = 2015, collapse_sex = T, collapse_age = my_age_groups)
```

## Exploring package data
### HRQoL norms - population-average health related quality of life data

For most countries, there is more than one set of HRQoL norms stored in the package data. For every country, we have chosen a set of default norms - these are the norms that will be used when the `calculate_dQALY` function is called without specifying a value for the `norms` argument. Alternatively, the user specify what set of package norms they would like to use by passing its ID to `norms`. Its also possible for the user to supply their own HRQoL population norm data.

The available HRQoL norms can be viewed using the `hrqol_norms` function. Results can be filtered by country, and returned with or without reference information.

```{r hrqol_norms}

# Return all utility norm sets for all countries
head(hrqol_norms())

# Return all English utility norm sets with reference information
head(hrqol_norms(country = "England", references = T))

```



```{r package_norm_plots}
# Calculating dQALY for four different countries using the country-specific 
# default utility norms set by the package
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T),
       aes(x = age, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   collapse_sex = T),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   collapse_sex = T),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   collapse_sex = T),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


# Calculating dQALY using the same set of utility norms for every country
# note that the ID used needs to be valid. e.g.:
# calculate_dQALY(country = "England", year = 2019, norms = "invalid") won't work
ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T,
                              norms = package_norms(country, id = "janssen_euvas")),
       aes(x = age, y = dQALY)) +
  geom_line(colour = "red") +
  geom_line(data = calculate_dQALY(country = "Argentina", year = 2019, 
                                   collapse_sex = T,
                                   norms = package_norms(country, id = "janssen_euvas")),
            colour = "green") +
  geom_line(data = calculate_dQALY(country = "France", year = 2019, 
                                   collapse_sex = T,
                                   norms = package_norms(country, id = "janssen_euvas")),
            colour = "blue") +
  geom_line(data = calculate_dQALY(country = "Hungary", year = 2019, 
                                   collapse_sex = T,
                                   #in hungary's case, EUVAS norms
                                   #are already the default
                                   norms = package_norms(country, id = "janssen_euvas")),
            colour = "purple") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


```


```{r custom_norms}
library(data.table)

# how to specify your own set of utility norms
my_norms <- data.table(sex = c(rep("male", 3), 
                               rep("female", 3)),
                       lower = c(0, 20, 90),
                       upper = c(19, 89, 150),
                       avg_hrqol = c(1, 0.85, 0.67,
                                    0.99, 0.4, 0.2))


ggplot(data = calculate_dQALY(country = "England", year = 2019,
                              collapse_sex = T),
       aes(x = age, y = dQALY)) +
  geom_line(colour = "black") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019,
                                   collapse_sex = T,
                                   norms = my_norms),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 30),
                     expand = c(0,0)) +
  theme_classic()


```

## Discounting
To get the net present value to society of the quality-adjusted life years that are 'lost' when an individual dies, we typically apply a discount rate which reflects the idea that, to us in the present, the value of QALYs that would be experienced years into the future is less than the value of QALYs experienced today. In our package, the default discount rate is set at 3.5% as per the [NICE health technology evaluations manual](https://www.nice.org.uk/process/pmg36/chapter/economic-evaluation-2#discounting). However, the most appropriate discount rate to apply will differ according to the context of the evaluation/analysis. [The Green Book](https://www.gov.uk/government/publications/the-green-book-appraisal-and-evaluation-in-central-government/the-green-book-2020#a6-discounting), guidance on evaluation methods issued by the Treasury, discusses a number of discounting regimes and the reasons one might use them. 

The function `calculate_dQALY` allows the user to specify the discount rate they would like to use in the calculation via setting the value of the argument ` r`. ` r` can be a scalar numeric (e.g. for a discount rate of 1%, set ` r = 0.01`) or, so that the discount rate can vary across time, ` r` also accepts vectorised functions (e.g. for a discount rate of 2% a year for 50 years into the future and 1% a year thereafter, set ` r = function(x) ifelse(x < 50, 0.02, 0.01)`). 

This package includes four discount rate functions that users can pass to `calculate_dQALY`, each of which implements a particular discounting regime recommended by the NICE HTA manual or Green Book:


<table>
    <tbody>
        <tr>
            <td>`r_default`</td>
            <td>NICE reference case discount rate/ Green Book standard Social Time Preference Rate</td>
        </tr>
        <tr>
            <td>`r_health`</td>
            <td>NICE alternative discount rate/Green Book recommended discount rate for health or life values</td>
        </tr>
        <tr>
            <td>`r_lt_health`</td>
            <td>Green Book recommended declining long term discount rate for health or life values</td>
        </tr>
        <tr>
            <td>`r_lt_health_reduced`</td>
            <td>Green Book recommended rate reduced by excluding pure social time preference (relevant if intervention may effect substantial/irreversible wealth transfers between generations)</td>
        </tr>
    </tbody>
</table>


The plots below show the values of the discount rate functions across time, and the impact of varying discount rates on the estimates of QALY loss associated with death across the life-course.

```{r discounting, echo = FALSE}

data.table(`Number of years into the future` = c(0:125),
                 r_default = r_default(c(0:125)),
                 r_health = r_health(c(0:125)),
                 r_lt_health = r_lt_health(c(0:125)),
                 r_lt_health_reduced = r_lt_health_reduced(c(0:125))) |> 
  melt(id.vars = "Number of years into the future",
       variable.name = "Discounting function",
       value.name = "Discount rate") |> 
  ggplot() +
  geom_line(aes(x = `Number of years into the future`, y = `Discount rate`, colour = `Discounting function`)) +
  scale_x_continuous(limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(-0.00013, 0.0355),
                     expand = c(0,0)) +
  theme_classic() +
  scale_colour_manual(values = c("red", "blue", "green", "orange"))



data.table(`Age at death` = c(0:104),
           r_default = calculate_dQALY(country = "England", year = 2019,
                                       collapse_sex = T,
                                       r = r_default)$dQALY,
           r_health = calculate_dQALY(country = "England", year = 2019,
                                       collapse_sex = T,
                                       r = r_health)$dQALY,
           r_lt_health = calculate_dQALY(country = "England", year = 2019,
                                       collapse_sex = T,
                                       r = r_lt_health)$dQALY,
           r_lt_health_reduced = calculate_dQALY(country = "England", year = 2019,
                                       collapse_sex = T,
                                       r = r_lt_health_reduced)$dQALY) |> 
  melt(id.vars = "Age at death",
       variable.name = "Discounting function",
       value.name = "QALY loss due to death") |> 
  ggplot() +
  geom_line(aes(x = `Age at death`, y = `QALY loss due to death`, colour = `Discounting function`)) +
  scale_x_continuous(limits = c(0, 105),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(0, 50),
                     expand = c(0,0)) +
  theme_classic() +
  scale_colour_manual(values = c("red", "blue", "green", "orange"))

```





## Note on methods
QALY loss at age x without discounting is equivalent to QALE at age x. QALY loss at age x without discounting or adjusting by health-related quality of life is life expectancy at age x.


```{r equivalencies}

# if we set the discount rate equal to zero, we'll be returning estimates of quality adjusted life expectancy
no_discounting <- 0

# if we ALSO set the health utility score for all ages equal to 1, we'll be returning life expectancy data
no_quality_adjustment <- data.table(sex = c("male", "female"),
                                    lower = 0,
                                    upper = 150,
                                    avg_hrqol = 1)


ggplot(data = calculate_dQALY(country = "England", year = 2019, 
                              collapse_sex = T), #default norms, default r
       aes(x = age, y = dQALY)) +
  geom_line() +
  geom_line(data = calculate_dQALY(country = "England", year = 2019,
                                   collapse_sex = T,
                                   r = no_discounting),
            colour = "blue") +
  # geom_line(data = calculate_QALE(country = "England", year = 2019,
  #                                   collapse_sex = T),
  #            colour = "green") +
  geom_line(data = calculate_dQALY(country = "England", year = 2019, 
                                   collapse_sex = T,
                                   r = no_discounting,
                                   norms = no_quality_adjustment),
            colour = "red") +
  scale_x_continuous(name = "Age at death",
                     limits = c(0, 125),
                     expand = c(0,0)) +
  scale_y_continuous(name = "QALY loss",
                     limits = c(0, 90),
                     expand = c(0,0)) +
  theme_classic()

```

## Worked example: using the dQALY package to value outputs of an infectious disease model
Imagine that we had built an infectious disease model, simulating CPE infections among hospital inpatients & the deaths resulting from those infections. Imagine that because of data availability, the model population does not have sex/age structure. We want to produce an estimate of the number of QALYs that would be lost on the death of an average hospital inpatient - this single value will be applied to all deaths. 

We'll use this example to demonstrate why one might use several of the features of the `calculate_dQALY` function - specifically, the features that enable the estimation of QALY losses for user-supplied cohorts and groups.


```{r worked_example_cohort}


# # Setting my cohort - all hospital inpatients
# # Using data on Hospital Admitted Patient Care Activity, 2019-20
# # https://digital.nhs.uk/data-and-information/publications/statistical/hospital-admitted-patient-care-activity/2019-20/summary-reports---apc---patient
# 
# temp <- tempfile(fileext = ".xlsx")
# download.file(url = "https://digital.nhs.uk/binaries/content/documents/corporate-website/publication-system/statistical/hospital-admitted-patient-care-activity/2019-20/summary-reports---apc---patient/summary-reports---apc---patient/publicationsystem%3AbodySections%5B2%5D/publicationsystem%3AdataFile",
#               temp, mode = "wb")
# 
# hospital_cohort <- as.data.table(readxl::read_xlsx(temp, sheet = 1)) |> 
#   setnames(new = c("x", "male", "female")) |> 
#   melt(measure.vars = c("male", "female"),
#        variable.name = "sex",
#        value.name = "count")
# 
# hospital_cohort <- hospital_cohort[x != "Unknown"][
#   , lower := as.numeric(substring(x, 1, 2))
# ][
#   , upper := lower + 4
# ][
#   , .(x = c(lower:upper), count = rep(count/5, 5)), by = c("lower", "upper", "sex")
# ][
#   , c("lower", "upper"):=NULL
# ]
# 
# 
# # Setting my age groups - want to group all ages together
# all_ages <- data.table(lower = 0, upper = 94)
# 
# 
# # Getting my QALY loss estimate
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = hospital_cohort)
# 
# 
# 
# # What if we were only modelling adult inpatients?
# adult_hospital_cohort <- hospital_cohort[x >= 18]
# 
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = adult_hospital_cohort)
# 
# 
# 
# # What if mortality rates were particularly high among newborns?
# adj_hospital_cohort <- copy(hospital_cohort)[x == 0, count := count*3]
# 
# calculate_dQALY(country = "England", year = 2020,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = adj_hospital_cohort)
# 
# 
# # What if we wanted to account for the fact that mortality and morbidity among
# # hospital patients is higher than it is among the general population?
# calculate_dQALY(country = "England", year = 2020,
#                 smr = 1.05, qcm = 0.95,
#                 collapse_age = all_ages,
#                 collapse_sex = TRUE,
#                 cohort = hospital_cohort)
# # Note: if you had your own life expectancy or health utility data for a hospital population
# # you could supply it directly to the function
# 
# 
# # Examining my estimates
# ggplot(data = calculate_dQALY(country = "England", year = 2020, 
#                               collapse_sex = T),
#        aes(x = age, y = dQALY)) +
#   geom_line(colour = "black") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = hospital_cohort)$dQALY, 
#             colour = "red",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = adult_hospital_cohort)$dQALY, 
#             colour = "blue",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = adj_hospital_cohort)$dQALY, 
#             colour = "green",
#             linetype = "dashed") +
#   geom_hline(yintercept = calculate_dQALY(country = "England", year = 2020,
#                                           smr = 1.05, qcm = 0.95,
#                                           collapse_age = all_ages,
#                                           collapse_sex = TRUE,
#                                           cohort = hospital_cohort)$dQALY, 
#             colour = "purple",
#             linetype = "dashed") +
#   scale_x_continuous(name = "Age at death",
#                      limits = c(0, 125),
#                      expand = c(0,0)) +
#   scale_y_continuous(name = "QALY loss",
#                      limits = c(0, 30),
#                      expand = c(0,0)) +
#   theme_classic()

```

