---
title: "Methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { equationNumbers: { autoNumber: "AMS" } }
});
</script>


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!-- ```{r setup} -->
<!-- library(dQALY) -->
<!-- ``` -->

We want to produce an estimate of $dQALY(x)$, the QALY loss associated with death at age $x$, for particular populations (usually country-level) and particular time periods (years). Our package implements a method for calculating QALY loss due to death described by Briggs et al. in a [letter](https://onlinelibrary.wiley.com/doi/10.1002/hec.4208) published in the journal Health Economics. 
<!-- This method requires two kinds of input data. -->

## Calculating QALY loss due to death
To produce estimates of QALY loss due to death for a given population, we require data regarding life expectancy and health-related quality of life within that population. 


Life tables report $q(x)$, the probability of dying between ages $x$ and $x+1$, for $x = 0,1,..,x_{max}$, $x_{max}$ being the max age we are interested in or have data for. With $q(x)$, we can calculate $l(x)$, the number surviving to age $x \ge 1$ for a reference population of size N,

$$l(x) = N\cdot\prod_{i=0}^{x}\big(1-q(i)\big)$$

and then (assuming a uniform distribution of death during the year) $L(x)$, the number of years lived between ages $x$ and $x+1$, 

$$L(x) = \frac{l(x) + l(x+1)}{2}$$

$L(x)$ and $l(x)$ can be used to calculate $LE(x)$, life expectancy at age $x$, as follows:

$$LE(x) = \frac{\sum_{i=x}^{x_{max}}L(i)}{l(x)}$$


The second kind of data needed is information about the average health-related quality of life (expressed as a number between 0 and 1) of people within the population at each age, which we can denote $Q(x)$. We'll refer to this type of data as HRQoL norms.

Life expectancy at age $x$ is the number of life years a person at age $x$ can expect to live over the remainder of their life - quality adjusted life expectancy, $QALE(x)$, is the number of quality-adjusted life years or QALYs that a person at age $x$ can expect to experience. It's given by

\begin{equation} \label{eq:qale}
QALE(x) = \frac{\sum_{i=x}^{x_{max}}Q(i)\cdot  L(i)}{l(x)} 
\end{equation}


<!-- There is one -->
When a person dies at age $x$ we don't say that the number of QALYs lost due to their death is exactly the same as the number of QALYs they were expected to experience over the remainder of their life - we apply a discount rate $r$ to the QALYs they would have accrued. This reflects the idea that to us in the present, the value of QALYs that would be experienced years into the future is less than the value of QALYs experienced today. The discount rate is usually set at 3.5%.

\begin{equation} \label{eq:dqaly}
dQALY(x) = \frac{\sum_{i=x}^{x_{max}}Q(i)\cdot  L(i)\cdot \frac{1}{(1+r)^{(i-x)}}}{l(x)}
\end{equation}

<br/>
<br/>
With the `dQALY` package, the functions `calculate_QALE` and `calculate_dQALY` implement the calculations described above. For more practical details on the use of these functions see the function documentation or the [demo vignette](demo.html).

<!-- Note that we follow Briggs et al. in using the terms $QALE$ to denote quality-adjusted life expectancy, and $dQALY$ to denote QALY loss due to death. We differ on this from McNamara.. -->


## Adjusting mortality and morbidity 

The method described above can be applied to any population for which we have life expectancy and health-related quality of life data. However, in practice, life tables and HRQoL norms are most commonly available at national level. If we would like to calculate QALY loss due to death in a population that we expect to have mortality and morbidity rates that are significantly different from the national average (e.g. people with Type I diabetes), then we can make adjustments to our calculation in order to reflect this difference.<br/>

A standardised mortality ratio $(SMR)$ is a measure that describes how mortality rates in a specific group differ with respect to some standard or reference population, after standardisation by age and sex. If the $SMR$ is greater than / lesser than 1, mortality rates in the group are greater than / lesser than rates in the reference population. <br/>

We want to use the estimates we have of the probability of dying between ages $x$ and $x+1$ in the reference population and the $SMR$ which captures the difference in mortality rates between the reference population and the group of interest to produce $q_{adj}(x)$, an adjusted probability of dying between ages $x$ and $x+1$.<br/>

If we let $q_{adj}(x) = SMR \cdot q(x)$, then it would be possible that where $x$ is large, $q_{adj}(x)$ would exceed 1. We need to covert $q(x)$, the probability of dying between ages $x$ and $x+1$, into the corresponding instantaneous death rate, which Briggs et al. denote $d(x)$, before applying the adjustment factor.<br/>

The relationship between $q(x)$ and $d(x)$ is the relationship between a probability and a rate:

$$q(x) =1 - e^{-d(x)} \iff d(x) = -ln\big(1-q(x)\big)$$ 

 
Instead of adjusting $q(x)$ directly, we apply our adjustment to $d(x)$, letting $d_{adj}(x) = SMR \cdot d(x)$ - so we can express $q_{adj}(x)$ in terms of $q(x)$ and $SMR$ as follows:<br/>


$$
\begin{aligned}
q_{adj}(x) &= 1 - e^{-d_{adj}(x)} \\&= 1 - e^{-SMR \cdot d(x)} \\&= 1 -  e^{-SMR \cdot \big( -ln(1-q(x))\big)} \\&= 1 - e^{ln((1-q(x))^{SMR})} \\&= 1 - (1 - q(x))^{SMR}
\end{aligned}
$$
So, 

$$
\begin{aligned}
l_{adj}(x) &= N \cdot \prod_{i=0}^{x}\big(1-q_{adj}(i)\big) \\&= N \cdot \prod_{i=0}^{x}\big(1-q(i)\big)^{SMR}
\end{aligned}
$$

and 

$$LE_{adj}(x) = \frac{l_{adj}(x) + l_{adj}(x+1)}{2}$$
<br/>
<br/>

Next, we'll use parameter $qCM$ to express our assumptions about how morbidity or health related quality of life in the population of interest differs from the reference population. With $qCM$, we try to quantify the impact of pre-existing comorbidities on quality of life. If $qCM$ is greater than / lesser than 1, then quality of life in the population of interest is better than / worse than quality of life in the reference population. We can adjust $Q(x)$, average quality of life in the reference population directly, letting 

$$Q_{adj}(x) = qCM \cdot Q(x)$$ 


From this point on, $QALE(x)$ and $dQALY(x)$ for the group of interest are derived from $l_{adj}(x)$, $LE_{adj}(x)$, and $Q_{adj}(x)$ exactly as shown previously in equations (1) and (2).<br/><br/>

In this package, the functions `calculate_QALE` and `calculate_dQALY` have arguments `smr` and `qcm`, which allow the user to adjust for mortality and morbidity as described. Again, see the function documentation or the [demo vignette](demo.html) for examples of this adjustment in practice.


## Note on deviations from the Briggs method
Our implementation of the method described in the Briggs letter differs in some minor respects to theirs, resulting in slightly different dQALY estimates. Briggs is calculating dQALY values for the United Kingdom using 2016-18 life tables - here are their estimates, taken from the Excel tool provided along with the letter, compared with the equivalent estimates generated by the `calculate_dQALY` function.

```{r}
library(data.table)
library(dQALY)

# reading from the Briggs excel tool 
temp <- tempfile(fileext = ".xlsx")
download.file(url = "https://www.lshtm.ac.uk/media/42556",
              temp, mode = "wb")

cbind(calculate_dQALY(country = "United Kingdom",
                      year = 2017,
                      collapse_sex = T,
                      collapse_age = data.frame(lower = seq(0, 90, 10),
                                                upper = seq(9, 99, 10)))[, c(1,4)],
      data.frame(briggs_dQALY = readxl::read_xlsx(temp, sheet = 2, range = "J17:J27")$dQALY))
```
The basic calculation remains the same - the differences are to do with choices around input data and grouping of estimates. In this section we'll show how to produce the Briggs results using the `calculate_dQALY` function - hopefully this exercise will help explain the source of the differences to those who are interested.  

### Life tables
Briggs is not interested in producing sex-specific estimates at any point. Instead of optionally pooling estimates at the end of the calculation he a) takes sex-specific q(x)’s, b) produces sex-specific l(x)’s, and then pools those l(x)’s together to get an average, and goes from there. We're just going to pull these values for l(x) in from the Briggs excel tool, and then - since we want to be able to use the `calculate_dQALY` function as its set up - we need the life expectancy information in terms of q(x), so we’ll back-calculate it:
```{r, eval = FALSE}
temp <- tempfile(fileext = ".xlsx")
download.file(url = "https://www.lshtm.ac.uk/media/42556",
              temp, mode = "wb")

lt <- as.data.table(readxl::read_xlsx(temp, sheet = 6, range = "R6:S127"))
lt[, q := 1 - shift(smrlx, type = "lead")/smrlx]
lt[, q := ifelse(is.na(q), shift(q, type = "lag"), q)]
lt <- lt[, .(age = Age, q)][, .(sex = c("male", "female")), by = .(age, q)]
```

### HRQoL norms
We can see in the excel tool (in the sheet called 'LookUpTables') that Briggs is using getting HRQoL norms for the UK from the Janssen paper, specifically the TTO values. This is data that we have stored in the package. 

We can also see that Briggs has assumed that the HRQoL norm for the youngest age group (0-17) is 1. The default assumption this package makes is more conservative - it takes the HRQoL norm for the 0-17 group from the next youngest group. To line up with Briggs, we can use the argument `avg_hrqol_young`:
```{r}
package_norms(country = "United Kingdom", 
              id = "janssen_tto",
              avg_hrqol_young = 1)
```

### Grouping estimates
In the sheet called 'Results' (which looks up values in sheet 'Calculations') we can see that Briggs is outputting estimates grouped into 10 year age bands. To produce a representative dQALY estimate for an age group, the excel tool selects the dQALY estimate belonging to the median age within the interval. 

The calculate_dQALY function does something different - it takes a weighted average of all the dQALY estimates in each group. By default, population-level age/sex distribution data provides the weightings. In order to mimic the Briggs output, in addition to defining 10 year age groups we need to construct a new cohort to use for the weighting which consists of one male/female aged 5, one male/female aged 15, and so on:
```{r}
age_grps <- data.frame(lower = seq(0, 90, 10),
                       upper = seq(9, 99, 10))
chrt <- data.frame(sex = c(rep("male", 10), rep("female", 10)),
                   age = c(seq(5, 95, 10), seq(5, 95, 10)),
                   count = 1)
```

### Matching Briggs output
The output of the following function call should match the Briggs results:
```{r, eval = F}
calculate_dQALY(life_table = lt,
                norms = package_norms(country = "United Kingdom", 
                                      id = "janssen_tto",
                                      avg_hrqol_young = 1),
                collapse_sex = T,
                collapse_age = age_grps,
                cohort = chrt)
```
